index

### 2024-03-15 appendix/index-definition.qmd
more index definitions

@@

@@

# spline

-   smooth
-   formula:
    -   $\hat{y}$ is derived from fitting a penalised spline model of y \~ x,
    -   $\hat{y}^\prime$ is derived from fitting a penalised spline model of x \~ y,
    -   the variance of the residuals is compared to the variance of the original data

$$I_{spline}(A) = \max \left(1 - \frac{var(y - \hat{y})}{var(x)}, 1 - \frac{var(y - \hat{y}^\prime)}{var(y)}\right)$$

    -   Ursula and Di (2020) Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics, page 1176: [link](https://arxiv.org/abs/1902.00181)
    -   code from the [cassowaryr page](https://github.com/numbats/cassowaryr/blob/HEAD/R/association.R)

>>>>

For 2D projections, we can also write $y_{n \times 2} = (y_1, y_2)$, where $y_1$ and $y_2$ are the two columns of the projected data.

@@

# dcor2d_2 index

-   smooth

-

-   Reference:

-   Note

  - From the function manual: "For bivariate data only, these are fast O(n log n) implementations of distance correlation and distance covariance statistics."

# MIC/ TIC index

-   smooth

-   formula

    -   Let $g(k, l)$ define a partition of the space $(y_1, y_2)$ into $k \times l$ rectangles, for example, $g(2, 3)$ means dividing the data space into 2 rectangles in the $y_2$ direction and 3 rectangles in the $y_1$ direction. Let $G$ denotes all the possible partition.
    -   MIC finds the maximum mutual information over G where k and l is bounded by the grid size: $k \times l < B(n) = n^{\alpha}$ where alpha = 0.3 in our simulation

    $$I_{MIC}(A) = \max_{g \in G} \frac{I(y_1, y_2 | g)}{\log(\min(k^*, l^*))}  = \max_{g \in G} \frac{\sum_{y1} \sum_{y2} P(y_1, y_2) \log \frac{P(y_1)}{P(y_1) P(y_2)}}{\log(\min(k^*, l^*))}$$

    where $k^*$ and $l^*$ are the number of rectangles in the optimal partition $g$.

    -   TIC calculates the sum of mutual information over G

$$I_{TIC}(A) = \sum_{g \in G} \frac{I(y_1, y_2 | g)}{\log(\min(k^*, l^*))}  = \sum_{g \in G} \frac{\sum_{y1} \sum_{y2} P(y_1, y_2) \log \frac{P(y_1)}{P(y_1) P(y_2)}}{\log(\min(k^*, l^*))}$$

-   Reference:

    -   Reshef (2011) Detecting Novel Associations in Large Datasets: [link](https://pubmed.ncbi.nlm.nih.gov/22174245/) - Figure 1 is usually to understand the general idea

    -   Reshef (2016) Measuring Dependence Powerfully and Equitably: [link](https://jmlr.csail.mit.edu/papers/volume17/15-308/15-308.pdf).

-   Note:

    -   The original Definition 1 on page 6 of Reshef (2016) uses supremum, here I just use the maximum.
    -   There are different version of MIC and we use $MIC_e$ and $TIC_e$ (See Sec 4 in the paper in Reshef (2016)), which uses the equicharacteristic matrix rather than the original characteristic matrix in both MIC/ TIC calculation, see page 14 Figure 1 for there difference (equicharacteristic matrix parts the space equally hence faster).
    -   Section 4.3/4.4 of Reshef (2016) talks about the time complexity

# loess/ spline index

-   smooth

-   formula:

    -   let $e^{\text{model}}_{y_1 \sim y_2}$ denote the residual from model $y_1 \sim y_2$

    -   for the loess index, we fit a loess model to $y_1 \sim y_2$ (use $\alpha = 0.05$ for the smoothing parameter in the loess index)

    $$I_{\text{loess}}(A) = \max \left(1 - \frac{var(e^{\text{loess}}_{y_1 \sim y_2})}{var(y_1)}, 1 - \frac{var(e^{\text{loess}}_{y_2 \sim y_1})}{var(y_2)}\right)$$

    -   for the spline index, a spline model is fitted (use cubic regression spline with k = 15 for the diemnsion of the basis in the spline model).

$$I_{\text{spline}}(A) = \max \left(1 - \frac{var(e^{\text{spline}}_{y_1 \sim y_2})}{var(y_1)}, 1 - \frac{var(e^{\text{spline}}_{y_2 \sim y_1})}{var(y_2)}\right)$$

-   reference:
    -   loess: (none) I create it myself, inspired by the spline index
    -   spline:
        -   Ursula and Di (2020) Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics, page 1176: [link](https://arxiv.org/abs/1902.00181)
        -   code from the [cassowaryr page](https://github.com/numbats/cassowaryr/blob/HEAD/R/association.R)

@@
